---
title: "Flow_analysis_10-30-17_miRFP-enrichment"
author: "haslen"
date: "October 17, 2017"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

This script uses flowCore to load data, then various functions to parse through it

## Load required libraries

```{r}
library(flowCore)
library(mclust)
library(ggplot2)
library(gridExtra)
library(grid)
library(lattice)
library(DescTools)
library(reshape2)
```

# Functions

## Clean data

Functions like log can cause -Inf values to occur - this function cleans thos up 

## trash gate

removes dead cell pieces from the FSC.SSC plot using simple filters

## Live cell filter

Now, filter dataset by fitting to bivariate normal distribution, picking mean of that distribution, and selecting all cells with the Mahalanobis distance, multiplied by the scale.factor argument.

Note that the relationship between Mahalanobis distance and probability is t = sqrt(-2*ln(1-p)), where t is the threshold mahalanobis disatnce and p is the propability.

So, for p = 0.001, the Mahalanobis distance is 3.72.

Note that the output of this is a filter, which needs to be applied to your dataset of interest, then the dataset of interest needs to be subset based on that filter

## Single cell filter

Now, filter by FSC.A by FSC.W using the same method

## Calculate ratio log(PE) / log(FITC) ratio (for Dendra2 experiments)

Now, calculate a ratio of activated to inactivated Dendra2 for each cell. 
Can't plot these values with FlowCore, so will just move into regular R dataframes

## mixed Gaussian modeling for datasets for which there is no negative control - can be used to separate two populations

```{r}
## make ggplot look good
ggplot <- function(...) { ggplot2::ggplot(...) +
                           theme(panel.background = element_rect(fill='white', colour='black'), # Wt bkg, blk border
                                 panel.grid.major = element_blank(),  # Hide major gridlines
                                 panel.grid.minor = element_blank(),  # Hide minor gridlines
                                 legend.key = element_rect(fill="white", size = 0.5))
  }


## Clean data
clean.data <- function(x){
  column_names <- colnames(x)
  x.clean <- x
  for(column in column_names) {
    x.clean <- x.clean[which(x.clean[,column] != '-Inf'),]
    x.clean <- x.clean[which(x.clean[,column] != 'NaN'),]
  }
  return(x.clean)
}

## trash gate
gate_trash = function(x, FSCA_cutoff, SSCA_cutoff) {
  x.notrash <- x[which(x[,'FSC.A'] > FSCA_cutoff & x[,'SSC.A'] > SSCA_cutoff),]
  return(x.notrash)
}

#x.notrash <- gate_trash(x, FSCA_cutoff = 15000, SSCA_cutoff = 15000)

## live cell gate

live.cell.gate = function(dat, p) {
  # calculate mahalanobis threshold
  mahalanobis_threshold = sqrt(-2*log(1-p))
  
  #set up filter using norm2filter
  n2f <- norm2Filter(x=c('FSC.A', 'SSC.A'), filterId='livecell',  scale.factor = mahalanobis_threshold)
  
  return(n2f)
}


## single cell gate

single.cell.gate = function(dat, p) {
  # calculate mahalanobis threshold
  mahalanobis_threshold = sqrt(-2*log(1-p))
  
  #set up filter using norm2filter
  n2f <- norm2Filter(x=c('FSC.A', 'FSC.W'), filterId='singlecell',  scale.factor = mahalanobis_threshold)
  
  return(n2f)
}

## calculate active:unactive Dendra2

calc.ratio.extract.data <- function(x){
  PE.A.log = log10(x@exprs[,'PE.A'])
  FITC.A.log = log10(x@exprs[,'FITC.A'])
  act.unact = PE.A.log/FITC.A.log
  A700.A.log   = log10(x@exprs[,'Alexa.Fluor.700.A'])
  new_data = as.data.frame(cbind(x@exprs, PE.A.log, FITC.A.log, A700.A.log, act.unact))
  new_data = clean.data(new_data)
  return(new_data)
}


##use mixed Gaussian to separate populations into two clusters (used for data for which there is not negative control)

onedim.gate.gausmix <- function(x, channel) {
  x.clean <- clean.data(x)
  # kmeans clustering on the desired channel with 2 clusters
  x.cluster <- Mclust(data.frame (x.clean[,channel]), G=2)
  # append cluster IDs to the cleaned dataframe
  x.clean.cluster <- cbind(x.clean, x.cluster$classification)
  # rename column so it doesn't have a $ sign in it
  cluster.name = paste0(channel, '.cluster')
  names(x.clean.cluster)[names(x.clean.cluster) == 'x.cluster$classification'] <- cluster.name
  # identify "positive" clusters using max function on channel in clean dataframe
  pos.cluster.id <- x.clean.cluster[which(x.clean.cluster[,channel] == max(x.clean.cluster[,channel])),cluster.name][1]
  # identify "negative" clusters using min function on channel in clean dataframe
  neg.cluster.id <- x.clean.cluster[which(x.clean.cluster[,channel] == min(x.clean.cluster[,channel])),cluster.name][1]
  # calculate mean of max value of neg and min value of pos and then return the cutoff value
  max.value.neg = max(x.clean.cluster[which(x.clean.cluster[,cluster.name] == neg.cluster.id),channel])
  min.value.pos = min(x.clean.cluster[which(x.clean.cluster[,cluster.name] == pos.cluster.id),channel])
  actunact.gate = mean(max.value.neg, min.value.pos)

  return(actunact.gate)
}

## label a hexbin plot corners with percentages for each gate, use with + in ggplot
hexbin_cutoff_label <- function(x, xvar, xgate, yvar, ygate) {
  # generate percentages
  perc_bottomleft <- round(100*length(x[which(x[,xvar] < xgate & x[,yvar] < ygate), xvar])/length(x[,xvar]),2)
  perc_bottomright <- round(100*length(x[which(x[,xvar] > xgate & x[,yvar] < ygate), xvar])/length(x[,xvar]),2)
  perc_topright <- round(100*length(x[which(x[,xvar] > xgate & x[,yvar] > ygate), xvar])/length(x[,xvar]),2)
  perc_topleft <- round(100*length(x[which(x[,xvar] < xgate & x[,yvar] > ygate), xvar])/length(x[,xvar]),2)
  # create labels
  xpos <- c(-Inf,-Inf,Inf,Inf)
  ypos <- c(-Inf, Inf,-Inf,Inf)
  annotateText <- c(paste0(perc_bottomleft, '%'),paste0(perc_topleft, '%'),paste0(perc_bottomright, '%'),paste0(perc_topright, '%'))
  hjustvar<-c(0,0,1,1) 
  vjustvar<-c(-1,1,-1,1)
  hexbin_cutoff_labels <- data.frame(xpos, ypos, hjustvar, vjustvar, annotateText)
  return(geom_text(data=hexbin_cutoff_labels, aes(label=annotateText, x=xpos, y=ypos, hjust = hjustvar, vjust = vjustvar)))
}

## return a dataframe of values of cutoffs using data and gates
cutoff_values_2D <- function(x, xvar, xgate, yvar, ygate) {
  # generate percentages
  xneg.yneg.count = length(x[which(x[,xvar] < xgate & x[,yvar] < ygate), xvar])
  xpos.yneg.count <- length(x[which(x[,xvar] > xgate & x[,yvar] < ygate), xvar])
  xpos.ypos.count <- length(x[which(x[,xvar] > xgate & x[,yvar] > ygate), xvar])
  xneg.ypos.count <- length(x[which(x[,xvar] < xgate & x[,yvar] > ygate), xvar])
  total <- length(x[,xvar])
  # create labels
  return(c(xneg.yneg.count, xpos.yneg.count, xpos.ypos.count, xneg.ypos.count, total))
}

## rank all values in the channel, then pick the cutoff closest to the the designated percentile
cutoff_value_ranked <- function(x, percentile) {
  sorted <- data.frame(values = sort(x, decreasing = FALSE, na.last = NA))
  sorted$percentile <- as.integer(row.names(sorted))/length(sorted$values)
  closest_value = which.min(abs(percentile - sorted$percentile))
  return(sorted[closest_value, 'values'])
}

# use kmeans clustering to fit a gate. Assumes 2 clusters
onedim.gate.kmeans <- function(x) {
  # kmeans clustering on the desired channel with 2 clusters
  x.cluster <- kmeans(x, 2, nstart = 100)
  # append cluster IDs to the cleaned dataframe
  x.annotated <- data.frame(cbind(x, x.cluster$cluster))
  colnames(x.annotated) <- c('values', 'cluster')
  # identify which cluster is high and which is low
  cluster.test <- data.frame('cluster' = str(0), 'meanvalue' = numeric(0))
  for(cluster in unique(x.annotated$cluster)) {
    cluster.mean = mean(x.annotated[which(x.annotated$cluster == cluster),'values'])
    cluster.test <- rbind(cluster.test, c(cluster, cluster.mean))
  colnames(cluster.test) <- c('cluster', 'mean')
  pos.cluster = cluster.test[which.max(cluster.test$mean), 'cluster']
  neg.cluster = cluster.test[which.min(cluster.test$mean), 'cluster']
  return(mean(min(x.annotated[which(x.annotated$cluster == pos.cluster),'values']), max(x.annotated[which(x.annotated$cluster == neg.cluster), 'values'])))
  }
  
  # identify "positive" clusters using max function on channel in clean dataframe
  pos.cluster.id <- x.clean.cluster[which(x.clean[,channel] == max(x.clean[,channel])),cluster.name]
  # identify "negative" clusters using min function on channel in clean dataframe
  neg.cluster.id <- x.clean.cluster[which(x.clean[,channel] == min(x.clean[,channel])),cluster.name]
  # 
  

  return(mean(pos))
}

# calculate sensitivity and specificity, given a dataframe with A700.A.log and act.unact + gates
sens.spec <- function(x, A700.gate, actunact.gate) {
  
  TP = length(which(x$A700.A.log > A700.gate & x$act.unact > actunact.gate))
  FP = length(which(x$A700.A.log < A700.gate & x$act.unact > actunact.gate))
  TN = length(which(x$A700.A.log < A700.gate & x$act.unact < actunact.gate))
  FN = length(which(x$A700.A.log > A700.gate & x$act.unact < actunact.gate))
  sensitivity = TP/(TP + FN)
  specificity = TN/(TN + FP)
  falsepositiverate = FP / (FP + TN)
  precision = TP/(TP+FP)
  return(data.frame(sens = sensitivity, spec = specificity, fpr = falsepositiverate, precision = precision))
}

```


## Run a control population through the pipeline

```{r}
# load data
data_name = 'U2OS C11 unactivated_A05_007'
# look at FSC.A and SSC.A
cntrl = read.FCS(paste0(data_name, '.fcs'), transformation=FALSE, alter.names=TRUE)
ggplot(as.data.frame(cntrl@exprs), aes(x=FSC.A, y=SSC.A, z=..density..)) + geom_hex() + scale_fill_distiller(palette = 'Spectral')

## gate out trash
trash = 25000
cntrl = read.FCS(paste0(data_name, '.fcs'), transformation=FALSE, alter.names=TRUE)
ggplot(as.data.frame(cntrl@exprs), aes(x=FSC.A, y=SSC.A, z=..density..)) + geom_hex() + scale_fill_distiller(palette = 'Spectral') + geom_vline(xintercept = trash) + geom_hline(yintercept = trash)
cntrl.notrash <- gate_trash(cntrl, FSCA_cutoff = trash, SSCA_cutoff = trash)

# generate live cell gate (called live) and check it
live = live.cell.gate(cntrl.notrash, 0.95)
# plot live gate
cntrl.live = Subset(cntrl.notrash, filter(cntrl.notrash, live))
ggplot(as.data.frame(cntrl.live@exprs), aes(x=FSC.A, y=SSC.A, z=..density..)) + geom_hex() + scale_fill_distiller(palette = 'Spectral') + scale_x_continuous(limits = c(min(cntrl@exprs[,'FSC.A']), max(cntrl@exprs[,'FSC.A']))) + scale_y_continuous(limits = c(min(cntrl@exprs[,'SSC.A']), max(cntrl@exprs[,'SSC.A'])))

# check FSC.A by FSC.W
ggplot(as.data.frame(cntrl.live@exprs), aes(x=FSC.A, y=FSC.W, z=..density..)) + geom_hex() + scale_fill_distiller(palette = 'Spectral')

# generate single cell gate and check it
single = single.cell.gate(cntrl.live, 0.999)
cntrl.live.single = Subset(cntrl.live, filter(cntrl.live, single))
ggplot(as.data.frame(cntrl.live.single@exprs), aes(x=FSC.A, y=FSC.W, z=..density..)) + geom_hex() + scale_fill_distiller(palette = 'Spectral') + scale_x_continuous(limits = c(min(cntrl.live@exprs[,'FSC.A']), max(cntrl.live@exprs[,'FSC.A']))) + scale_y_continuous(limits = c(min(cntrl.live@exprs[,'FSC.W']), max(cntrl.live@exprs[,'FSC.W'])))

# extract data, calculate PE/FITC ratio
cntrl.live.single <- clean.data(calc.ratio.extract.data(cntrl.live.single))

# use gaussian mixture model to gate out FITC
FITC.gate <- onedim.gate.gausmix(cntrl.live.single, 'FITC.A.log')
cntrl.live.single.FITCpos<-cntrl.live.single[which(cntrl.live.single[,'FITC.A.log'] > FITC.gate),]
ggplot(cntrl.live.single, aes(x = FITC.A.log)) + geom_histogram(bins = 50) + geom_vline(xintercept = FITC.gate, linetype = 'dashed', size = 1.5)

# plot activated/unactivated dendra2 ratio
ggplot(cntrl.live.single.FITCpos, aes(x = act.unact)) + geom_histogram(bins = 75)

# use a ranked metric to pick actunact gate (gates out 99% of cells)
actunact.gate = cutoff_value_ranked(cntrl.live.single.FITCpos$act.unact, 0.995)
ggplot(cntrl.live.single.FITCpos, aes(x = act.unact)) + geom_histogram(bins = 50) + geom_vline(xintercept = actunact.gate, linetype = 'dashed', size = 1.5)

# plot miRFP +ve cells
ggplot(cntrl.live.single.FITCpos, aes(x = A700.A.log)) + geom_histogram(bins = 50) + geom_vline(xintercept = actunact.gate, linetype = 'dashed', size = 1.5)

# set an A700 gate
A700_gate = 3
ggplot(cntrl.live.single.FITCpos, aes(x = A700.A.log)) + geom_histogram(bins = 50) + geom_vline(xintercept = A700_gate, linetype = 'dashed', size = 1.5)


# plot of miRFP vs. act.unact with labels in place
ggplot(cntrl.live.single.FITCpos, aes(x=act.unact, y=A700.A.log)) + geom_hex(bins = 50) + scale_fill_distiller(palette = 'Spectral') + hexbin_cutoff_label(cntrl.live.single.FITCpos, 'act.unact', actunact.gate, 'A700.A.log', A700_gate) + geom_hline(yintercept = A700_gate, linetype = 'dashed') + geom_vline(xintercept = actunact.gate, linetype = 'dashed')



```

## analyze data for ROC curves

```{r}
# plot ROC curves

# collect names of all data in directory and remove .fcs from files
file_list = list.files(path = '.', pattern = '.fcs')
file_list_nofcs <- unlist(strsplit(file_list, split = '.fcs'))

metadat <- data.frame(file_list_nofcs, c(T, T, T, T, T, T, T, T, T, T, T, T, T, F, F, F, F, F, F, F, F, F, F, F), factor(c(10, 50, 0.4, 0.4, 2, 10, 50, 2, 0.4, 50, 10, 2, 0.4, 100, 100, 50, 10, 2, 0.4, 50, 0.4, 0.4, 0, 50)), stringsAsFactors = FALSE)

colnames(metadat) <- c('sample', 'activated', 'expected.prop.miRFP')


i <- 1

AUC_values = c()
prop_miRFPpos = c()
sens = c()
spec = c()
precision = c()
fpr = c()
plot_data = list()

for(file in file_list) {
  # prep plot list
  plots = list()
  
  # convert data name to file
  data_name = strsplit(file, split = '.fcs')
  
  # read data
  x = read.FCS(paste0(data_name, '.fcs'), transformation=FALSE, alter.names=TRUE)
  
  ## gate out trash
  x.notrash <- gate_trash(x, FSCA_cutoff = 25000, SSCA_cutoff = 25000)
  
  # filter for live cell and plot
  x.live = Subset(x.notrash, filter(x.notrash, live))
  
  # filter for single cell
  x.single = Subset(x.live, filter(x.live, single))
  
  # extract data and calculate PE/FITC ratio
  x.live.single <- calc.ratio.extract.data(x.single)
  # remove rows with data at the minimum value for PE.A, FITC.A
  x.live.single <- x.live.single[which(x.live.single[,'PE.A.log'] != min(x.live.single[,'PE.A.log'])),]
  x.live.single <- x.live.single[which(x.live.single[,'FITC.A.log'] != min(x.live.single[,'FITC.A.log'])),]
  
  # filter for FITC positive cells
  x.live.single.FITCpos<-x.live.single[which(x.live.single[,'FITC.A.log'] > FITC.gate),]
  
  plots[[length(plots) + 1]] = ggplot(x.live.single, aes(x = FITC.A.log)) + geom_histogram(bins = 75) + geom_vline(xintercept = FITC.gate, linetype = 'dashed', size = 1)
  
  # calculate the actual % positive miRFP cells
  prop_miRFPpos[i] = length(which(x.live.single.FITCpos$A700.A.log > A700_gate))/length(x.live.single.FITCpos$A700.A.log)
  
  # calculate the sensitivity and specificity at baseline
  sens.spec.data <- sens.spec(x.live.single.FITCpos, A700_gate, actunact.gate)
  sens[i] <- sens.spec.data$sens
  spec[i] <- sens.spec.data$spec
  fpr[i] <- sens.spec.data$fpr
  precision[i] <- sens.spec.data$precision
  
  # calculate ROC and AUC
  quants_poscells = quantile(x.live.single.FITCpos[which(x.live.single.FITCpos$act.unact > actunact.gate),]$act.unact, probs = seq(0, 1, 0.01), names = FALSE)
  quants_poscells = quantile(x.live.single.FITCpos$act.unact, probs = seq(0, 1, 0.01), names = FALSE)
  quants_poscells = data.frame(seq(0, 1, 0.01), quants_poscells)
  colnames(quants_poscells) = c('percentile','cutoff')
  ROCdata <- cbind(quants_poscells, as.data.frame(t(as.data.frame(sapply(quants_poscells$cutoff, sens.spec, x = x.live.single.FITCpos, A700.gate = A700_gate)))))
  ROCdata$sens = unlist(ROCdata$sens)
  ROCdata$spec = unlist(ROCdata$spec)
  ROCdata$fpr = unlist(ROCdata$fpr)
  ROCdata <- ROCdata[order(ROCdata$sens),]
  plots[[length(plots) + 1]] = ggplot(ROCdata, aes(x = (1 - spec), y= sens)) + geom_point() + geom_line()
  xAUC = 1 - (ROCdata$spec)
  yAUC = ROCdata$sens
  AUC_values[i] <- AUC(xAUC, yAUC, method = c("spline"))
  
  # plot 2D plot of PE.A vs. FITC
  plots[[length(plots) + 1]] = ggplot(x.live.single.FITCpos, aes(x=FITC.A.log, y=PE.A.log)) + geom_hex(bins = 50) + scale_fill_distiller(palette = 'Spectral')
  
  # plot 1D plot of act.unact
  plots[[length(plots) + 1]] = ggplot(x.live.single.FITCpos, aes(x = act.unact)) + geom_histogram(bins = 75)
  
  # plot the 2D hexbin of act.unact vs. A.700.A.log
  plots[[length(plots) + 1]] = ggplot(x.live.single.FITCpos, aes(x=act.unact, y=A700.A.log)) + geom_hex(bins = 50) + scale_fill_distiller(palette = 'Spectral') + hexbin_cutoff_label(x.live.single.FITCpos, 'act.unact', actunact.gate, 'A700.A.log', A700_gate) + geom_hline(yintercept = A700_gate, linetype = 'dashed') + geom_vline(xintercept = actunact.gate, linetype = 'dashed')
  
  # plot the 2D hexbin showing all bins with 1 cell in it as grey
  plots[[length(plots) + 1]] = ggplot(x.live.single.FITCpos, aes(x=act.unact, y=A700.A.log)) + geom_hex(bins = 50) + scale_fill_distiller(palette = 'Spectral', limits = c(2,350)) + hexbin_cutoff_label(x.live.single.FITCpos, 'act.unact', actunact.gate, 'A700.A.log', A700_gate) + geom_hline(yintercept = A700_gate, linetype = 'dashed') + geom_vline(xintercept = actunact.gate, linetype = 'dashed')
  
  # place plots into directory in a single pdf
  #pdf_name = paste0(data_name, '_plots.pdf')
  #pdf(pdf_name, onefile = TRUE)
  #grid.arrange(grobs=plots, ncol=2, top=str(data_name))
  #dev.off()
  
  i <- i +1
  
  # return plotting data
  plot_data[data_name[[1]]] = x.live.single.FITCpos
}

metadat <- cbind(metadat, prop_miRFPpos, AUC_values, spec, sens, fpr, precision)

metadat$expected.prop.miRFP <- as.factor(metadat$expected.prop.miRFP)

to_plot <- melt(metadat[which(metadat$activated),c(1:4,8,9)], id.vars = c('sample', 'activated', 'expected.prop.miRFP', 'prop_miRFPpos'))

ggplot(to_plot, aes(x = prop_miRFPpos, y = value, color = variable)) + geom_point() #+ stat_summary(geom = 'line', aes(x = expected.prop.miRFP, y = value), fun.y = mean)

ggplot(metadat[which(metadat$activated == TRUE),], aes(x = expected.prop.miRFP, y = AUC_values, group = expected.prop.miRFP)) + geom_boxplot() + ylim(0.5, 1) 

ggplot(metadat[which(metadat$activated == TRUE),], aes(x = expected.prop.miRFP, y = precision, group = expected.prop.miRFP)) + geom_boxplot()

ggplot(metadat[which(metadat$activated == TRUE),], aes(x = expected.prop.miRFP, y = sens, group = expected.prop.miRFP)) + geom_boxplot()

ggplot(metadat[which(metadat$activated == TRUE & metadat$sample != 'U2OS activated_B03_010'),], aes(x = 1-spec, y = sens, label = round(spec, digits = 3), color = expected.prop.miRFP)) + geom_point(size = 3) + geom_text(vjust = -0.5)

write.csv(metadat, 'miRFPenrich_metadat_09-18-17.csv')


```

