---
title: "Activation Accuracy"
author: "Nick Hasle"
date: "April 16, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(flowCore)
library(mclust)
library(ggplot2)
library(gridExtra)
library(grid)
library(lattice)
library(DescTools)
library(reshape2)

## scientific notation
fancy_scientific <- function(l) {
     # turn in to character string in scientific notation
     l <- format(l, scientific = TRUE)
     # quote the part before the exponent to keep all the digits
     l <- gsub("^(.*)e", "'\\1'e", l)
     # turn the 'e+' into plotmath format
     l <- gsub("e", "%*%10^", l)
     # return this as an expression
     parse(text=l)
}

```

# Functions

## Clean data

Functions like log can cause -Inf values to occur - this function cleans thos up 

## trash gate

removes dead cell pieces from the FSC.SSC plot using simple filters

## Live cell filter

Now, filter dataset by fitting to bivariate normal distribution, picking mean of that distribution, and selecting all cells with the Mahalanobis distance, multiplied by the scale.factor argument.

Note that the relationship between Mahalanobis distance and probability is t = sqrt(-2*ln(1-p)), where t is the threshold mahalanobis disatnce and p is the propability.

So, for p = 0.001, the Mahalanobis distance is 3.72.

Note that the output of this is a filter, which needs to be applied to your dataset of interest, then the dataset of interest needs to be subset based on that filter

## Single cell filter

Now, filter by FSC.A by FSC.W using the same method

## Calculate ratio log(PE) / log(FITC) ratio (for Dendra2 experiments)

Now, calculate a ratio of activated to inactivated Dendra2 for each cell. 
Can't plot these values with FlowCore, so will just move into regular R dataframes


```{r}
## Clean data
clean.data <- function(x){
  column_names <- colnames(x)
  x.clean <- x
  for(column in column_names) {
    x.clean <- x.clean[which(x.clean[,column] != '-Inf'),]
    x.clean <- x.clean[which(x.clean[,column] != 'NaN'),]
  }
  return(x.clean)
}

## trash gate
gate_trash = function(x, FSCA_cutoff, SSCA_cutoff) {
  x.notrash <- x[which(x[,'FSC.A'] > FSCA_cutoff & x[,'SSC.A'] > SSCA_cutoff),]
  return(x.notrash)
}

#x.notrash <- gate_trash(x, FSCA_cutoff = 15000, SSCA_cutoff = 15000)

## live cell gate

live.cell.gate = function(dat, p) {
  # calculate mahalanobis threshold
  mahalanobis_threshold = sqrt(-2*log(1-p))
  
  #set up filter using norm2filter
  n2f <- norm2Filter(x=c('FSC.A', 'SSC.A'), filterId='livecell',  scale.factor = mahalanobis_threshold)
  
  return(n2f)
}


## single cell gate

single.cell.gate = function(dat, p) {
  # calculate mahalanobis threshold
  mahalanobis_threshold = sqrt(-2*log(1-p))
  
  #set up filter using norm2filter
  n2f <- norm2Filter(x=c('FSC.A', 'FSC.W'), filterId='singlecell',  scale.factor = mahalanobis_threshold)
  
  return(n2f)
}

## calculate active:unactive Dendra2

calc.ratio.extract.data <- function(x){
  PE.A.log = log10(x@exprs[,'PE.A'])
  FITC.A.log = log10(x@exprs[,'FITC.A'])
  act.unact = PE.A.log/FITC.A.log
  A700.A.log   = log10(x@exprs[,'Alexa.Fluor.700.A'])
  new_data = as.data.frame(cbind(x@exprs, PE.A.log, FITC.A.log, A700.A.log, act.unact))
  new_data = clean.data(new_data)
  return(new_data)
}


## label a hexbin plot corners with percentages for each gate, use with + in ggplot
hexbin_cutoff_label <- function(x, xvar, xgate, yvar, ygate) {
  # generate percentages
  perc_bottomleft <- round(100*length(x[which(x[,xvar] < xgate & x[,yvar] < ygate), xvar])/length(x[,xvar]),2)
  perc_bottomright <- round(100*length(x[which(x[,xvar] > xgate & x[,yvar] < ygate), xvar])/length(x[,xvar]),2)
  perc_topright <- round(100*length(x[which(x[,xvar] > xgate & x[,yvar] > ygate), xvar])/length(x[,xvar]),2)
  perc_topleft <- round(100*length(x[which(x[,xvar] < xgate & x[,yvar] > ygate), xvar])/length(x[,xvar]),2)
  # create labels
  xpos <- c(-Inf,-Inf,Inf,Inf)
  ypos <- c(-Inf, Inf,-Inf,Inf)
  annotateText <- c(paste0(perc_bottomleft, '%'),paste0(perc_topleft, '%'),paste0(perc_bottomright, '%'),paste0(perc_topright, '%'))
  hjustvar<-c(0,0,1,1) 
  vjustvar<-c(-1,1,-1,1)
  hexbin_cutoff_labels <- data.frame(xpos, ypos, hjustvar, vjustvar, annotateText)
  return(geom_text(data=hexbin_cutoff_labels, aes(label=annotateText, x=xpos, y=ypos, hjust = hjustvar, vjust = vjustvar)))
}

## return a dataframe of values of cutoffs using data and gates
cutoff_values_2D <- function(x, xvar, xgate, yvar, ygate) {
  # generate percentages
  xneg.yneg.count = length(x[which(x[,xvar] < xgate & x[,yvar] < ygate), xvar])
  xpos.yneg.count <- length(x[which(x[,xvar] > xgate & x[,yvar] < ygate), xvar])
  xpos.ypos.count <- length(x[which(x[,xvar] > xgate & x[,yvar] > ygate), xvar])
  xneg.ypos.count <- length(x[which(x[,xvar] < xgate & x[,yvar] > ygate), xvar])
  total <- length(x[,xvar])
  # create labels
  return(c(xneg.yneg.count, xpos.yneg.count, xpos.ypos.count, xneg.ypos.count, total))
}

## rank all values in the channel, then pick the cutoff closest to the the designated percentile
cutoff_value_ranked <- function(x, percentile) {
  sorted <- data.frame(values = sort(x, decreasing = FALSE, na.last = NA))
  sorted$percentile <- as.integer(row.names(sorted))/length(sorted$values)
  closest_value = which.min(abs(percentile - sorted$percentile))
  return(sorted[closest_value, 'values'])
}


# calculate sensitivity and specificity, given a dataframe with A700.A.log and act.unact + gates
sens.spec <- function(x, A700.gate, actunact.gate) {
  
  TP = length(which(x$A700.A.log > A700.gate & x$act.unact > actunact.gate))
  FP = length(which(x$A700.A.log < A700.gate & x$act.unact > actunact.gate))
  TN = length(which(x$A700.A.log < A700.gate & x$act.unact < actunact.gate))
  FN = length(which(x$A700.A.log > A700.gate & x$act.unact < actunact.gate))
  sensitivity = TP/(TP + FN)
  specificity = TN/(TN + FP)
  falsepositiverate = FP / (FP + TN)
  precision = TP/(TP+FP)
  return(data.frame(sens = sensitivity, spec = specificity, fpr = falsepositiverate, precision = precision))
}

##use mixed Gaussian to separate populations into two clusters (used for data for which there is not negative control)

onedim.gate.gausmix <- function(x, channel) {
  x.clean <- clean.data(x)
  # kmeans clustering on the desired channel with 2 clusters
  x.cluster <- Mclust(data.frame (x.clean[,channel]), G=2)
  # append cluster IDs to the cleaned dataframe
  x.clean.cluster <- cbind(x.clean, x.cluster$classification)
  # rename column so it doesn't have a $ sign in it
  cluster.name = paste0(channel, '.cluster')
  names(x.clean.cluster)[names(x.clean.cluster) == 'x.cluster$classification'] <- cluster.name
  # identify "positive" clusters using max function on channel in clean dataframe
  pos.cluster.id <- x.clean.cluster[which(x.clean.cluster[,channel] == max(x.clean.cluster[,channel])),cluster.name][1]
  # identify "negative" clusters using min function on channel in clean dataframe
  neg.cluster.id <- x.clean.cluster[which(x.clean.cluster[,channel] == min(x.clean.cluster[,channel])),cluster.name][1]
  # calculate mean of max value of neg and min value of pos and then return the cutoff value
  max.value.neg = max(x.clean.cluster[which(x.clean.cluster[,cluster.name] == neg.cluster.id),channel])
  min.value.pos = min(x.clean.cluster[which(x.clean.cluster[,cluster.name] == pos.cluster.id),channel])
  actunact.gate = mean(max.value.neg, min.value.pos)

  return(actunact.gate)
}


```


## Figure 1A: Run a control sample through pipeline

```{r}
# load data
data_name = '../data/U2OS C11 unactivated_A05_007'
# look at FSC.A and SSC.A
cntrl = read.FCS(paste0(data_name, '.fcs'), transformation=FALSE, alter.names=TRUE)
ggplot(as.data.frame(cntrl@exprs), aes(x=FSC.A, y=SSC.A, z=..density..)) + geom_hex() + scale_fill_distiller(palette = 'Spectral')

## gate out trash
trash = 25000
cntrl = read.FCS(paste0(data_name, '.fcs'), transformation=FALSE, alter.names=TRUE)
ggplot(as.data.frame(cntrl@exprs), aes(x=FSC.A, y=SSC.A, z=..density..)) + geom_hex() + scale_fill_distiller(palette = 'Spectral') + geom_vline(xintercept = trash) + geom_hline(yintercept = trash)
cntrl.notrash <- gate_trash(cntrl, FSCA_cutoff = trash, SSCA_cutoff = trash)

# generate live cell gate (called live) and check it
live = live.cell.gate(cntrl.notrash, 0.95)
# plot live gate
cntrl.live = Subset(cntrl.notrash, filter(cntrl.notrash, live))
ggplot(as.data.frame(cntrl.live@exprs), aes(x=FSC.A, y=SSC.A, z=..density..)) + geom_hex() + scale_fill_distiller(palette = 'Spectral') + scale_x_continuous(limits = c(min(cntrl@exprs[,'FSC.A']), max(cntrl@exprs[,'FSC.A']))) + scale_y_continuous(limits = c(min(cntrl@exprs[,'SSC.A']), max(cntrl@exprs[,'SSC.A'])))

# check FSC.A by FSC.W
ggplot(as.data.frame(cntrl.live@exprs), aes(x=FSC.A, y=FSC.W, z=..density..)) + geom_hex() + scale_fill_distiller(palette = 'Spectral')

# generate single cell gate and check it
single = single.cell.gate(cntrl.live, 0.999)
cntrl.live.single = Subset(cntrl.live, filter(cntrl.live, single))
ggplot(as.data.frame(cntrl.live.single@exprs), aes(x=FSC.A, y=FSC.W, z=..density..)) + geom_hex() + scale_fill_distiller(palette = 'Spectral') + scale_x_continuous(limits = c(min(cntrl.live@exprs[,'FSC.A']), max(cntrl.live@exprs[,'FSC.A']))) + scale_y_continuous(limits = c(min(cntrl.live@exprs[,'FSC.W']), max(cntrl.live@exprs[,'FSC.W'])))

# extract data, calculate PE/FITC ratio
cntrl.live.single <- clean.data(calc.ratio.extract.data(cntrl.live.single))

# use gaussian mixture model to gate out FITC
FITC.gate <- onedim.gate.gausmix(cntrl.live.single, 'FITC.A.log')
cntrl.live.single.FITCpos<-cntrl.live.single[which(cntrl.live.single[,'FITC.A.log'] > FITC.gate),]
ggplot(cntrl.live.single, aes(x = FITC.A.log)) + geom_histogram(bins = 50) + geom_vline(xintercept = FITC.gate, linetype = 'dashed', size = 1.5)

# plot activated/unactivated dendra2 ratio
ggplot(cntrl.live.single.FITCpos, aes(x = act.unact)) + geom_histogram(bins = 75)

# use a ranked metric to pick actunact gate (gates out 99% of cells)
actunact.gate = 0.85
ggplot(cntrl.live.single.FITCpos, aes(x = act.unact)) + geom_histogram(bins = 50) + geom_vline(xintercept = actunact.gate, linetype = 'dashed', size = 1.5)

# plot miRFP +ve cells
ggplot(cntrl.live.single.FITCpos, aes(x = A700.A.log)) + geom_histogram(bins = 50) + geom_vline(xintercept = actunact.gate, linetype = 'dashed', size = 1.5)

# set an A700 gate
A700_gate = 3
ggplot(cntrl.live.single.FITCpos, aes(x = A700.A.log)) + geom_histogram(bins = 50) + geom_vline(xintercept = A700_gate, linetype = 'dashed', size = 1.5)


# plot of miRFP vs. act.unact with labels in place
ggplot(cntrl.live.single.FITCpos, aes(x=act.unact, y=A700.A.log)) + geom_hex(bins = 50) + scale_fill_distiller(palette = 'Spectral') + hexbin_cutoff_label(cntrl.live.single.FITCpos, 'act.unact', actunact.gate, 'A700.A.log', A700_gate) + geom_hline(yintercept = A700_gate, linetype = 'dashed') + geom_vline(xintercept = actunact.gate, linetype = 'dashed')



```


## Run all data and analyze ROC curves

```{r}
# plot ROC curves

# collect names of all data in directory and remove .fcs from files
file_list = list.files(path = '../data/', pattern = '.fcs')
file_list_nofcs <- unlist(strsplit(file_list, split = '.fcs'))

metadat <- read.csv('metadat.csv')


# initialize values of interest
{i <- 1 # iterate through sample list; initialize iteration at 1
AUC_values = c()
prop_miRFPpos = c()
sens = c()
spec = c()
precision = c()
fpr = c()
flow_plot_data = list()

flow_plot_data = lapply(file_list, function(file) {
  # prep plot list
  plots = list()
  
  # convert data name to file
  data_name = strsplit(file, split = '.fcs')
  
  # read data
  x = read.FCS(paste0('../data/', data_name, '.fcs'), transformation=FALSE, alter.names=TRUE)
  
  ## gate out trash
  x.notrash <- gate_trash(x, FSCA_cutoff = 25000, SSCA_cutoff = 25000)
  
  # filter for live cell and plot
  x.live = Subset(x.notrash, filter(x.notrash, live))
  
  # filter for single cell
  x.single = Subset(x.live, filter(x.live, single))
  
  # extract data and calculate PE/FITC ratio
  x.live.single <- calc.ratio.extract.data(x.single)
  # remove rows with data at the minimum value for PE.A, FITC.A
  x.live.single <- x.live.single[which(x.live.single[,'PE.A.log'] != min(x.live.single[,'PE.A.log'])),]
  x.live.single <- x.live.single[which(x.live.single[,'FITC.A.log'] != min(x.live.single[,'FITC.A.log'])),]
  
  # filter for FITC positive cells
  x.live.single.FITCpos<-x.live.single[which(x.live.single[,'FITC.A.log'] > FITC.gate),]
  
  plots[[length(plots) + 1]] = ggplot(x.live.single, aes(x = FITC.A.log)) + geom_histogram(bins = 75) + geom_vline(xintercept = FITC.gate, linetype = 'dashed', size = 1)
  
  # calculate the actual % positive miRFP cells
  prop_miRFPpos[i] <<- length(which(x.live.single.FITCpos$A700.A.log > A700_gate))/length(x.live.single.FITCpos$A700.A.log)
  
  # calculate the sensitivity and specificity at baseline
  sens.spec.data <- sens.spec(x.live.single.FITCpos, A700_gate, actunact.gate)
  sens[i] <<- sens.spec.data$sens
  spec[i] <<- sens.spec.data$spec
  fpr[i] <<- sens.spec.data$fpr
  precision[i] <<- sens.spec.data$precision
  
  # calculate ROC and AUC
  quants_poscells = quantile(x.live.single.FITCpos[which(x.live.single.FITCpos$act.unact > actunact.gate),]$act.unact, probs = seq(0, 1, 0.01), names = FALSE)
  quants_poscells = quantile(x.live.single.FITCpos$act.unact, probs = seq(0, 1, 0.01), names = FALSE)
  quants_poscells = data.frame(seq(0, 1, 0.01), quants_poscells)
  colnames(quants_poscells) = c('percentile','cutoff')
  ROCdata <- cbind(quants_poscells, as.data.frame(t(as.data.frame(sapply(quants_poscells$cutoff, sens.spec, x = x.live.single.FITCpos, A700.gate = A700_gate)))))
  ROCdata$sens = unlist(ROCdata$sens)
  ROCdata$spec = unlist(ROCdata$spec)
  ROCdata$precision = unlist(ROCdata$precision)
  ROCdata$fpr = unlist(ROCdata$fpr)
  ROCdata <- ROCdata[order(ROCdata$sens),]
  plots[[length(plots) + 1]] = ggplot(ROCdata, aes(x = (1 - precision), y= sens)) + geom_point() + geom_line()
  xAUC = 1 - (ROCdata$precision)
  yAUC = ROCdata$sens
  AUC_values[i] <<- AUC(xAUC[which(!is.na(xAUC))], yAUC[which(!is.na(xAUC))], method = c("spline"))
  
  # plot 2D plot of PE.A vs. FITC
  plots[[length(plots) + 1]] = ggplot(x.live.single.FITCpos, aes(x=FITC.A.log, y=PE.A.log)) + geom_hex(bins = 50) + scale_fill_distiller(palette = 'Spectral')
  
  # plot 1D plot of act.unact
  plots[[length(plots) + 1]] = ggplot(x.live.single.FITCpos, aes(x = act.unact)) + geom_histogram(bins = 75) + geom_vline(xintercept = actunact.gate, linetype = 'dashed')
  
  # plot the 2D hexbin of act.unact vs. A.700.A.log
  plots[[length(plots) + 1]] = ggplot(x.live.single.FITCpos, aes(x=act.unact, y=A700.A.log)) + geom_hex(bins = 50) + scale_fill_distiller(palette = 'Spectral') + hexbin_cutoff_label(x.live.single.FITCpos, 'act.unact', actunact.gate, 'A700.A.log', A700_gate) + geom_hline(yintercept = A700_gate, linetype = 'dashed') + geom_vline(xintercept = actunact.gate, linetype = 'dashed')
  
  # plot the 2D hexbin showing all bins with 1 cell in it as grey
  plots[[length(plots) + 1]] = ggplot(x.live.single.FITCpos, aes(x=act.unact, y=A700.A.log)) + geom_hex(bins = 50) + scale_fill_distiller(palette = 'Spectral', limits = c(2,350)) + hexbin_cutoff_label(x.live.single.FITCpos, 'act.unact', actunact.gate, 'A700.A.log', A700_gate) + geom_hline(yintercept = A700_gate, linetype = 'dashed') + geom_vline(xintercept = actunact.gate, linetype = 'dashed')

  i <<- i +1
  
  # return plotting data
  flow_plot_data[data_name[[1]]] = x.live.single.FITCpos
  
  # return the AUC curve
  return(x.live.single.FITCpos)
})

metadat <- cbind(metadat, prop_miRFPpos, AUC_values, spec, sens, fpr, precision)

metadat$expected.prop.miRFP <- as.factor(metadat$expected.prop.miRFP)}


```

Generate Figure 1B

```{r}
library(dplyr)
actunact.gate = 0.85
A700_gate = 3

# take two 10% datasets
datasets = flow_plot_data[which(metadat$sample %in% c('U2OS C11 unactivated_A03_005', 'U2OS activated_C04_018'))]

# merge the datasets together
datasets[[1]]$label = factor('Activated Culture', levels = c('Unactivated\nControl', 'Activated Culture'))
datasets[[2]]$label = 'Unactivated\nControl'
datasets = bind_rows(datasets)

# get the labels for percentages
datasets$activated = F
datasets[which(datasets$act.unact > actunact.gate),'activated'] = T
datasets$miRFPpos = F
datasets[which(datasets$A700.A.log > A700_gate),'miRFPpos'] = T

# get % positive for each
print(100*table(datasets[which(datasets$label == 'Unactivated\nControl'),c('miRFPpos','activated')])/sum((datasets$label == 'Unactivated\nControl')))

print(100*table(datasets[which(datasets$label == 'Activated Culture'),c('miRFPpos','activated')])/sum((datasets$label == 'Activated Culture')))

# plot the datasets
ylablist = expression(Log["10"]*'(miRFP Intensity)')
xlablist = expression(atop(Log["10"]*'(Activated/Unactivated', 'Dendra2 Intensity)'))
#ylablist = 'miRFP Intensity'
datasets$label = factor(datasets$label, levels = c('Activated Culture', 'Unactivated\nControl'), ordered = T)

# get density for psuedocolor plot
get_density <- function(x, y, ...) {
  dens <- MASS::kde2d(x, y, ...)
  ix <- findInterval(x, dens$x)
  iy <- findInterval(y, dens$y)
  ii <- cbind(ix, iy)
  return(dens$z[ii])
}

datasets$density = get_density(datasets$act.unact, datasets$A700.A.log, n =100)

ggplot(datasets, aes(x = act.unact, y = A700.A.log, color = density)) + 
  geom_point(size = 0.1, alpha = 0.5) +
  #geom_hex(bins = 50, color = 'black', size = 0.1) + 
  scale_fill_distiller(limits = c(2,550)) + 
  geom_hline(yintercept = A700_gate, color = 'black') + 
  geom_vline(xintercept = actunact.gate, color = 'black') + 
  xlab(xlablist) + ylab(ylablist) +
  xlim(0,1.3) +
  ylim(-0.5, 5.5) +
  facet_wrap(~label, ncol = 2, scales = 'free_y') + 
  guides(color = FALSE) + 
  theme_classic() +
  theme(axis.text=element_text(size=12), 
        axis.title=element_text(size=14), 
        strip.text.x = element_text(size=14), 
        strip.background = element_blank()) +
  scale_colour_gradientn(colours = rev(c('#ffffd9','#edf8b1','#c7e9b4','#7fcdbb','#41b6c4','#1d91c0','#225ea8','#253494','#081d58'))) +
  guides(color = FALSE)
  
ggsave('F1B.png', width = 4.25, height = 3.25, dpi = 900)

```

Generate figure 1C

```{r}
# calculate average precision
mean(metadat[which(metadat$activated == T),'precision'])
mean(metadat[which(metadat$activated == T),'sens'])

to_plot <- melt(metadat[which(metadat$activated),c(1:4,7,9)], id.vars = c('sample', 'activated', 'expected.prop.miRFP', 'prop_miRFPpos'))
to_plot$expected.prop.miRFP = as.numeric(as.character(to_plot$expected.prop.miRFP))/100

ggplot(to_plot, aes(x = expected.prop.miRFP, 
                    y = value, 
                    color = variable)) + 
  geom_point(alpha = 0.5) + 
  stat_summary(geom = 'line', 
               aes(x = expected.prop.miRFP, 
                   y = value, 
                   group = variable), 
               fun.y = mean, size = 1) + 
  stat_summary(fun.data = 'mean_cl_boot', 
               aes(x = expected.prop.miRFP, 
                   y = value, 
                   group = variable), 
               size = 0.5) + 
  theme_classic() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     limits = c(0.5,1)) + 
  scale_color_manual(values = c('#d7191c', '#2c7bb6')) + 
  ylab('Value') + 
  xlab('Proportion miRFP-positive\nCells in Coculture') + 
  guides(color = FALSE) + 
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14),
        axis.title.y = element_blank())

ggsave('F1C.png', width = 3.3, height = 3, dpi = 900)
```


## load data for S1A

```{r}
data = read.csv('../data/Dendra2Decay_S1A/all_data_filtered_annotated.csv')

head(data)
```

## Plot S1A

```{r}
ggplot(data[which(data$activation_ms != 0),], aes(x = ratio, color = as.factor(timepoint))) + 
  geom_density(data = data[which(data$activation_ms == 0),c('ratio', 'doxycycline')], color = 'grey') +
  geom_density() +
  geom_classic() +
  facet_grid(activation_ms~doxycycline) +
  scale_color_manual(values = rev(c('#fcbba1','#fc9272','#fb6a4a','#de2d26','#a50f15'))) +
  xlab('Activated / Unactivated Dendra2') +
  guides(color=guide_legend(title="Time Since Activation (hr)")) +
  xlim(-8,0.25)

ggsave('Dendra2_Decay_S1A.png', dpi = 900, width = 7.5, height = 4)
  
```

## load data for S1B

```{r}
data = read.csv('../data/405nm-toxicity_S1B/FloJo-analysis/01-26-18_data.csv', stringsAsFactors = F)

data = melt(data[1:18,], id.vars = c('X', 'activation', 'Replicate'))

# remove data for activated cells in negative/positive control (these cells were never activated) and data for unactivated cells in the activated wells (we are interested in the cell death in activated cells only)
## column for whether cells were activated
data$activated = sapply(data$variable, function(s) unlist(strsplit(as.character(s), '\\.'))[3])
## column for subpop
data$subpop = sapply(data$variable, function(s) unlist(strsplit(as.character(s), '\\.'))[4])
## remove any cells which dont EITHER have 
data = data[which(((data$activation %in% c('negative_control', 'positive_control') & (data$activated == 'unactivated_cells')) | ((data$activation %in% c('2h', '8h', '24h', '48h')) & (data$activated == 'activated_cells'))) & (data$subpop %in% c('apop', 'dead', 'necr'))),]
## make activation ordered factor
data[which(data$activation == 'negative_control'),'activation'] = 'NegC'
data[which(data$activation == 'positive_control'),'activation'] = 'PosC'
data$activation = factor(data$activation, levels = c('NegC', 'PosC', '2h', '8h', '24h', '48h'))

# relabel the apop etc.
data$subpop <- factor(data$subpop, levels=c("apop", "necr", "dead"), labels=c("Apoptotic", "Necrotic", "Dead"))

head(data)
```

## plot S1A (gating scheme)

```{r}
  # read data
  x = read.FCS('../data/U2OS activated_C04_018.fcs', transformation=FALSE, alter.names=TRUE)
  x.toplot = as.data.frame(x@exprs)
  x.toplot$density = get_density(x.toplot$FSC.A, x.toplot$SSC.A, n = 100)
  x.plt = ggplot(x.toplot, aes(x=FSC.A, y=SSC.A, color = density)) +
    theme_classic() +
    geom_point(size = 0.5) +
    scale_colour_gradientn(colours = rev(c('#ffffd9','#edf8b1','#c7e9b4','#7fcdbb','#41b6c4','#1d91c0','#225ea8','#253494','#081d58'))) +
    geom_vline(xintercept =25000, linetype='dashed') +
    geom_hline(yintercept =25000, linetype='dashed') +
    guides(color = F)

  ## gate out trash
  x.notrash <- gate_trash(x, FSCA_cutoff = 25000, SSCA_cutoff = 25000)
  x.notrash.toplot = as.data.frame(x.notrash@exprs)
  x.notrash.toplot$density = get_density(x.notrash.toplot$FSC.A, x.notrash.toplot$SSC.A, n = 100)
  x.notrash.plt = ggplot(x.notrash.toplot, aes(x=FSC.A, y=SSC.A, color = density)) +
    theme_classic() +
    geom_point(size = 0.5) +
    scale_colour_gradientn(colours = rev(c('#ffffd9','#edf8b1','#c7e9b4','#7fcdbb','#41b6c4','#1d91c0','#225ea8','#253494','#081d58'))) +
    guides(color = F)
  
  # filter for live cell and plot
  x.live = Subset(x.notrash, filter(x.notrash, live))
  x.live.toplot = as.data.frame(x.live@exprs)
  x.live.toplot$density = get_density(x.live.toplot$FSC.A, x.live.toplot$SSC.A, n = 100)
  x.live.plt = ggplot(x.live.toplot, aes(x=FSC.A, y=SSC.A, color = density)) +
    geom_point(data = x.notrash.toplot, alpha = 0) +
    theme_classic() +
    geom_point(size = 0.5) +
    scale_colour_gradientn(colours = rev(c('#ffffd9','#edf8b1','#c7e9b4','#7fcdbb','#41b6c4','#1d91c0','#225ea8','#253494','#081d58')))+
    guides(color = F)
  
  # filter for single
  x.live.toplot$density2 = get_density(x.live.toplot$FSC.A, x.live.toplot$FSC.W, n = 100)
  x.live.plt2 = ggplot(x.live.toplot, aes(x=FSC.A, y=FSC.W, color = density2)) +
    #geom_point(data = x.notrash.toplot, alpha = 0) +
    theme_classic() +
    theme(axis.text.x = element_text(angle =45, hjust=1 ))+
    geom_point(size = 0.5) +
    scale_colour_gradientn(colours = rev(c('#ffffd9','#edf8b1','#c7e9b4','#7fcdbb','#41b6c4','#1d91c0','#225ea8','#253494','#081d58')))+
    guides(color = F) + scale_y_continuous(labels = function(x) format(x, scientific = TRUE)) +
    scale_x_continuous(labels = function(x) format(x, scientific = TRUE))
  
  # filter for single cell
  x.single = Subset(x.live, filter(x.live, single))
  x.single.toplot = as.data.frame(x.single@exprs)
  x.single.toplot$density = get_density(x.single.toplot$FSC.A, x.single.toplot$FSC.W, n = 100)
  x.single.plt = ggplot(x.single.toplot, aes(x=FSC.A, y=FSC.W, color = density)) +
    geom_point(data = x.live.toplot, alpha = 0) +
    theme_classic() +
    theme(axis.text.x = element_text(angle =45, hjust=1 ))+
    geom_point(size = 0.5) +
    scale_colour_gradientn(colours = rev(c('#ffffd9','#edf8b1','#c7e9b4','#7fcdbb','#41b6c4','#1d91c0','#225ea8','#253494','#081d58')))+
    guides(color = F) +
    scale_y_continuous(labels = function(x) format(x, scientific = TRUE)) +
    scale_x_continuous(labels = function(x) format(x, scientific = TRUE)) 
  
  # extract data and calculate PE/FITC ratio
  x.live.single <- calc.ratio.extract.data(x.single)
  # remove rows with data at the minimum value for PE.A, FITC.A
  x.live.single <- x.live.single[which(x.live.single[,'PE.A.log'] != min(x.live.single[,'PE.A.log'])),]
  x.live.single <- x.live.single[which(x.live.single[,'FITC.A.log'] != min(x.live.single[,'FITC.A.log'])),]
  
  # show dendra2 filter
  x.live.singleplt = ggplot(x.live.single, aes(x=FITC.A.log)) +
    geom_histogram(bins = 100) +
    theme_classic() +
    geom_vline(xintercept = FITC.gate, linetype = 'dashed')
  
  # filter for FITC positive cells
  x.live.single.FITCpos<-x.live.single[which(x.live.single[,'FITC.A.log'] > FITC.gate),]

S1A = ggarrange(plotlist = list(x.plt, x.notrash.plt, x.live.plt, x.live.plt2, x.single.plt, x.live.singleplt), ncol = 3, nrow = 2)

ggsave('SF1A.png', S1A, width = 7.5, height = 4.5, dpi = 900)
```


## plot S1C

```{r}


ggplot(data, aes(x = activation, y = value/100, color = subpop)) + 
  geom_point(alpha = 0.3, size = 2) +
  ylab('Percentage') +
  xlab('Condition') +
  guides(color=guide_legend(title="Cell Viability")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))
  

ggsave('S1C.png', dpi = 900, width = 4, height = 2.75)
```

